master thesis

### 2018-03-21 Paper - Latex/Paper - Skizze - v4.tex
Paper - Introduction

%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%--------------------------------------------------------------------------------------------------- 
\chapter{Introduction}
  This paper is about the analysis of ancestral state reconstruction algorithms for non-binary trees, 
    applied to the currently largest phylogeny synthesis tree of Open Tree Of Life, with the 
    application of prediciton of parasitism. \\
  \todo{Or:} The aim of this thesis is the application of maximum parsiomony algorithms to non-binary 
    trees and very large datasets. In particular, the example 'Origins of Parasitism' throughout the 
    Eukaryota Tree of Life. \\
  We will test the existing algorithms Sankoff (R package \textit{Castor}\cite{Louca2017}) and Fitch 
    for this task and estimate their predictive power. In addition, we will extend the Fitch 
    algorithm from binary trees to multifurcating trees to make it usable and comparable for our 
    current problem. \\
  This achieves that we can predict ancestral states and unknown states of living species for large 
    non-binary relatives trees. \\

  For about 50 years, people have been working on ancestral state reconstruction. The first paper 
    mentioned above is by Camin and Sokal, who in 1965 were working on algorithms for discrete-state 
    data \cite{Camin1965}. Different methods have been developed and the question is which method is 
    the most suitable for our problem. \\

  \todotext{In den Bacteria und Archaea wurde das mal allgemein für binary traits gemacht: \\
  "A total 90~\% of all binary traits described molecular functions, specifically the presence or absence of a gene or a set of gene involved in biochemical pathways."
  \cite{Goberna2015} \\
  Unterschied bei uns ist parasitismus komplexer.. da sind meist nicht einzelne Gene zuständig..} \\

  For this purpose, various studied methods and their advantages and disadvantages will be compared. \\
  These algorithms were developed especially for the binary problem, since one has considered much 
    smaller subtrees, of which one knows all the splits. \\
  By developing a whole of an entire tree of life, the problem arises that it is by no means binary.

  % Researchers of the phylogenies have been dealt with the ancenstral state reconstruction in the 
  %   60s. The first methods were only brute force \todo{Quelle, siehe Fitch: Camin and Sokal 1965}. 
  %   Next came a set of parsimony algorithms such as: Fitch-parsimony \cite{Fitch1971}, 
  %   Wagner-parsimony \cite{Swofford1987} ... \todo{weitere?}. \\
  %   With more and more data, there is now the possibility to use more information to calculate the 
  %   probabilities of the ancestral states. In addition to the states of the leafs, algorithms could 
  %   also use branch lengths. The likelihood based algorithms came more in interest. \\
  %   Our focus came with another 'data extension'. We wanted to work with the biggest phylogenetic tree 
  %   that exists at this moment, which goes over all observed species. For most \todo{most?} species 
  %   there is no phylogeny, but only a taxonomic classification.
  To accomplish this task, a large phylogenetic tree and information about the current species states 
    is needed. %In this work we use the synthesis tree of Open Tree of Life (OTL) \cite{Hinchliff2015} 
  The biggest 'phylogenetic tree' is a synthesis of phylogenetic trees filled with a taxonomic tree 
    given by Open Tree of Life \cite{Hinchliff2015}. This tree is not binary and therefore the 
    developed algorithms are not directly applicable.Therefore, an algorithm must be found that can 
    work with non-binary trees. \\
  For the information about the current states of the species we use the interaction database Global 
    Biotic Interactions (GloBI) \cite{Poelen2014}.

  % In this work, we have looked at the algorithms that are generally suited to our data, to develop 
  % them further for the not binary case, and finally to compare their usability with our sythesis 
  % tree. \\
  % We have decided to consider only parsimony algorithms since we have no information on branch 
  % lengths and no other additional information like different transition probabilities of our states.

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %--------------------------------------------------------------------------------------------------- 
  \section{Definitions}

    \begin{itemize}
      \item Multifurkation - binär (binary or bifurcating... \cite{Felsenstein2003})
      \item height (min, max, mean), depth of a tree/node (Distanz zur Wurzel vs distanz zum Blatt)
      \item OTL, OTT, GloBI
    \end{itemize}

    %---------------------------------------------------------------------------------------------------
    %--------------------------------------------------------------------------------------------------- 
    \subsection{Parasite}
      Since we use Globi to classify species as parasitic or free-living, we use their definition of 
        parasitism. In GloBi, ontobee definitions are used.
      The interaction has parasite is defined as: "An interaction relationship between two organisms 
        living together in more or less intimate association in a relationship in which association is 
        disadvantageous or destructive to one of the organisms." \cite{...}. This definition includes: ecto,-
      and endoparasites, parasitoids, kleptoparasites and pathogenes.

      \todotext{Mungall, C., (2017). Definition for the interaction-term: "parasitised by; has parasite" ob
      ontobee.org. Last checked: 24.07.2017 at 
      \hyperlink{ontobee.org/ontology/RO?iri=http://purl.obolibrary.org/obo/RO_0002445}
      {ontobee.org/ontology/...}}

    %---------------------------------------------------------------------------------------------------
    %--------------------------------------------------------------------------------------------------- 
    \subsection{Maximum parsimony}
      Maximum parsimony methods Maximum parsimony in phylogeny refers to favoring the tree that needs 
        the least evolutionary change to explain the observed data. \\
      In our case, it is about the change of states 'is free-living' or 'is parasitic'.

@@

      \begin{table}
        \begin{center}
          \hspace*{-2cm}\begin{tabular}{ |l|r||r|r||r|r|r|r|r|r|r|r| }
            \hline
            & & \multicolumn{2}{c||}{original states} & \multicolumn{8}{c|}{final states} \\
            Kingdom & \# nodes & FL & P
              & 0 (FL) & 0.25 & 0.33 & 0.4 & 0.5 & 0.67 & 0.75 & 1 (P) \\
            \hline \hline
            none & 75446 & 45 & 529 
              & 13426 & 220 & 24082 & 0 & 7792 & 5302 & 0 & 24493 \\
            Fungi & 31457 & 577 & 2983
              & 38520 & 0 & 0 & 0 & 5723 & 0 & 0 & 266463 \\
            Chloroplastida & 416478 & 3519 & 77
              & 410795 & 0 & 0 & 0 & 4182 & 0 & 0 & 1501 \\
            Metazoa & 1491012 & 30758 & 22373
              & 1328135 & 0 & 0 & 930 & 25535 & 4423 & 1665 & 130324 \\
            \hline  
          \end{tabular}
        \end{center}
        \caption{Kingdom (leaf nodes)}
        \label{table:Kingdom leaf nodes}
      \end{table}

      \begin{table}
        \begin{center}
          \hspace*{-2cm}\begin{tabular}{ |l|r||r|r||r|r|r|r|r|r|r|r| }
            \hline
            & & \multicolumn{2}{c||}{original states} & \multicolumn{8}{c|}{final states} \\
            Kingdom & \# nodes & FL & P
              & 0 (FL) & 0.25 & 0.33 & 0.4 & 0.5 & 0.67 & 0.75 & 1 (P) \\
            \hline \hline
            none & 84456 & 45 & 529 
              & 15035 & 243 & 25910 & 0 & 8764 & 6183 & 0 & 28140 \\
            Fungi & 324105 & 577 & 2983
              & 39088 & 0 & 0 & 0 & 5858 & 0 & 0 & 274803 \\
            Chloroplastida & 460457 & 3519 & 77
              & 454211 & 0 & 0 & 0 & 4688 & 0 & 0 & 1558 \\
            Metazoa & 1670956 & 30758 & 22373
              & 1485749 & 0 & 0 & 1313 & 29002 & 5102 & 1957 & 147833 \\
            \hline  
          \end{tabular}
        \end{center}
        \caption{Kingdom (inkl internal nodes)}
        \label{table:Kingdom internal nodes}
      \end{table}

@@

      \begin{lstlisting}[gobble=6]
        # possible tags: 0, 0.333, 0.4, 0.5, 0.667, 0.75, 1
        # rounded to:    0  0      0    0    1      1     1
        if node_state != father_state:
          if father_state == 0:
              origins += 1        # FL -> P
              new_found = True
          else:
              losses += 1         # P -> FL
              new_found = True
      \end{lstlisting}

      \todo{without rounding change else: to elif $father_state == 1$}

>>>>

%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%--------------------------------------------------------------------------------------------------- 
\chapter{Introduction}
  This paper is about the analysis of ancestral state reconstruction algorithms for non-binary trees, 
    applied to the currently largest phylogenetic synthesis tree of Open Tree Of Life (OTL)
    \cite{Hinchliff2015}, with the application of prediciton of parasitism. \\
  % \todo{Or:} The aim of this paper is the application of maximum parsiomony algorithms to non-binary 
  %   trees and very large datasets. In particular, the example 'Origins of Parasitism' throughout the 
  %   Eukaryota Tree of Life. \\

  We will test the existing maximum parsimony algorithms Fitch \cite{Fitch1971} and Sankoff 
    \cite{Sankoff1975} for this task and estimate their predictive power. The present tree of OTL is 
    not binary. A tree is binary (or bifurcated), when every node has at most two children otherwise 
    it is multifurcated \cite{Felsenstein2003}. \\
  Maximum parsimony methods are developed for phylogenies, which are usually depicted as binary trees.
    Parsimonious in phylogeny refers to favoring the tree that needs the least evolutionary change 
    to explain the observed data. In our case, it is about the change of states 'is free-living' or 
    'is parasitic'. \\
    % Parsimonious in this context means to minimize the number of transitions between species states 
    % (free-living or parasitic). 
  The Sankoff method is implemented by Louca et al. for the non-binary case and is available as an R 
    package called \textit{Castor} \cite{Louca2017}. In addition, we have implemented the Fitch 
    method and adapted it for multifurcated trees. \\% to make it usable and comparable for our current problem.
  This achieves that we can predict ancestral states and unknown states of living species for large 
    non-binary relatives trees. \\

  For about 50 years, people have been working on ancestral state reconstruction. The first paper 
    is by Camin and Sokal, who in 1965 were working on algorithms for discrete-state data 
    \cite{Camin1965}. Different methods have been developed and the question is which method is the 
    most suitable for our problem. \\
  % The first methods were only brute force \todo{Quelle, siehe Fitch: Camin and Sokal 1965}. 
  %   Next came a set of parsimony algorithms such as: Fitch-parsimony \cite{Fitch1971}, 
  %   Wagner-parsimony \cite{Swofford1987} ... \todo{weitere?}. \\
  %   With more and more data, there is now the possibility to use more information to calculate the 
  %   probabilities of the ancestral states. In addition to the states of the leafs, algorithms could 
  %   also use branch lengths. The likelihood based algorithms came more in interest. \\

  % We have decided to consider only parsimony algorithms since we have no information on branch 
  % lengths and no other additional information like different transition probabilities of our states.

  For this purpose, various studied methods and their advantages and disadvantages will be compared. \\
  These algorithms were developed especially for the binary problem, since one has considered much 
    smaller subtrees, of which one knows all the splits. \\
  By developing a whole of an entire tree of life, the problem arises that it is by no means binary. \\

  For these large phylogenetic synthesis trees, however, ancestral state reconstruction has so far 
    only been done for Bacteria and Archaea for binary traits by Goberna and Verdú \cite{Goberna2015}.
    However, this differs from eukaryotes in the sense that complex traits such as parasitism depend
    on more than one gene. \\

  To accomplish this task, a large phylogenetic tree and information about the current species states 
    is needed. \\
  The biggest 'phylogenetic tree' is a synthesis of phylogenetic trees filled with a taxonomic trees 
    given by Open Tree of Life (OTL) \cite{Hinchliff2015}. %This tree is not binary and therefore the 
    % developed algorithms are not directly applicable. Therefore, an algorithm must be found that can 
    % work with non-binary trees. \\
  For the information about the current states of the species we use the interaction database Global 
    Biotic Interactions (GloBI) \cite{Poelen2014}. The data in GloBI are stored as interactions. 
    Useful for us e.g. Species A parasitize species B. From this we conclude that species A is 
    parasitic and species B free-living. In the case parasite parasitizes parasite, we get conflicting 
    states for the second species. We solve this by preferring parasitic. \\

  At this point a few words to the term parasitic. There are different definitions. Since we use Globi 
    to classify species, we use their definition of parasitism. Again, in GloBi, ontobee definitions 
    are used.
  The interaction has parasite is defined as: "An interaction relationship between two organisms 
    living together in more or less intimate association in a relationship in which association is 
    disadvantageous or destructive to one of the organisms." \cite{...}. This definition includes: 
    ecto- and endoparasites, parasitoids, kleptoparasites and pathogenes.
  \todotext{Mungall, C., (2017). Definition for the interaction-term: "parasitised by; has parasite" ob
  ontobee.org. Last checked: 24.07.2017 at 
  \hyperlink{ontobee.org/ontology/RO?iri=http://purl.obolibrary.org/obo/RO_0002445}
  {ontobee.org/ontology/...}} \\

  This paper thus consists of the following points: (1) Find a suitable ancestral state 
    reconstruction method. (2) Perform reconstructing on the Eucaryotic synthesis tree of OTL. \\
  Point 1 consists of a simulation of our data situation and the application of various possible 
    algorithms to these. To evaluate them based on the results. \\
  The one by Louca et al. implemented Sankoff algorithm is the best in our comparisons. Therefore, 
    point 2 consists of reconstructing the ancestral states and predicting the unknown leaf states. 
    And then an evaluation of the results. \\

  \todotext{Definitions of height (min, max, mean), depth of a tree/node (Distanz zur Wurzel vs 
    distanz zum Blatt)}

@@

      \begin{table}[h!]
        \begin{center}
          \hspace*{-2cm}\begin{tabular}{ |l|r||r|r||r|r|r|r|r|r|r|r| }
            \hline
            & & \multicolumn{2}{c||}{original states} & \multicolumn{8}{c|}{final states} \\
            Kingdom & \# nodes & FL & P
              & 0 (FL) & 0.25 & 0.33 & 0.4 & 0.5 & 0.67 & 0.75 & 1 (P) \\
            \hline \hline
            none & 75446 & 45 & 529 
              & 13426 & 220 & 24082 & 0 & 7792 & 5302 & 0 & 24493 \\
            Fungi & 31457 & 577 & 2983
              & 38520 & 0 & 0 & 0 & 5723 & 0 & 0 & 266463 \\
            Chloroplastida & 416478 & 3519 & 77
              & 410795 & 0 & 0 & 0 & 4182 & 0 & 0 & 1501 \\
            Metazoa & 1491012 & 30758 & 22373
              & 1328135 & 0 & 0 & 930 & 25535 & 4423 & 1665 & 130324 \\
            \hline  
          \end{tabular}
        \end{center}
        \caption{Kingdom (leaf nodes)}
        \label{table:Kingdom leaf nodes}
      \end{table}

      \begin{table}[h!]
        \begin{center}
          \hspace*{-2cm}\begin{tabular}{ |l|r||r|r||r|r|r|r|r|r|r|r| }
            \hline
            & & \multicolumn{2}{c||}{original states} & \multicolumn{8}{c|}{final states} \\
            Kingdom & \# nodes & FL & P
              & 0 (FL) & 0.25 & 0.33 & 0.4 & 0.5 & 0.67 & 0.75 & 1 (P) \\
            \hline \hline
            none & 84456 & 45 & 529 
              & 15035 & 243 & 25910 & 0 & 8764 & 6183 & 0 & 28140 \\
            Fungi & 324105 & 577 & 2983
              & 39088 & 0 & 0 & 0 & 5858 & 0 & 0 & 274803 \\
            Chloroplastida & 460457 & 3519 & 77
              & 454211 & 0 & 0 & 0 & 4688 & 0 & 0 & 1558 \\
            Metazoa & 1670956 & 30758 & 22373
              & 1485749 & 0 & 0 & 1313 & 29002 & 5102 & 1957 & 147833 \\
            \hline  
          \end{tabular}
        \end{center}
        \caption{Kingdom (inkl internal nodes)}
        \label{table:Kingdom internal nodes}
      \end{table}

@@

      \begin{lstlisting}[gobble=6]
        possible tags: 0, 0.333, 0.4, 0.5, 0.667, 0.75, 1
        rounded to:    0  0      0    0    1      1     1
      \end{lstlisting}

      % \begin{lstlisting}[gobble=6]
      %   # possible tags: 0, 0.333, 0.4, 0.5, 0.667, 0.75, 1
      %   # rounded to:    0  0      0    0    1      1     1
      %   if node_state != father_state:
      %     if father_state == 0:
      %         origins += 1        # FL -> P
      %         new_found = True
      %     else:
      %         losses += 1         # P -> FL
      %         new_found = True
      % \end{lstlisting}

      % \todo{without rounding change else: to elif $father_state == 1$}
